\savegeometry{fhsststandard}
\newgeometry{paper=a4paper, lmargin=0.1\paperwidth, rmargin=0.1\paperwidth, tmargin=1in, bmargin=1in}


\begin{titlepage}
\begin{adjustwidth}{1in}{1in}
\begin{center}
    \thispagestyle{empty}

    \vspace*{4in}

    {\normalfont\sffamily\fontsize{36}\normalfont\itshape{Everything Maths } \\ \vspace*{1cm}
    {\normalfont\sffamily\fontsize{22}\normalfont\itshape{Grade 10 Mathematics}}
    \vspace*{1in} \\
    \LARGE Version 1 -- CAPS \\

   {\vspace*{2in}
     by Siyavula and volunteers 
  

\vfill

    }}
\end{center}
\end{adjustwidth}
\end{titlepage}





% Authors
\newpage
\thispagestyle{empty}


\begin{flushleft} \textbf{\LARGE Authors List} \end{flushleft}

{This book is based upon the original Free High School Science Text which was entirely written by
volunteer academics, educators and industry professionals. Their vision was to see a curriculum aligned
set of mathematics and physical science textbooks which are freely available to anybody and exist
under an open copyright license.} \par

\textbf{\large Siyavula core team} \\

Neels van der Westhuizen; Alison Jenkin; Leonard Gumani Mudau; Marina van Zyl; Helen Robertson; Carl Scheffler; Nicola du Toit; Josephine Mamaroke Phatlane; William Buthane Chauke; Thomas Masango \par

\textbf{\large Original Free High School Science Texts core team}\\

Mark Horner; Samuel Halliday; Sarah Blyth; Rory Adams; Spencer Wheaton \par 


\textbf{\large Original Free High School Science Texts editors}\\

Jaynie Padayachee; Joanne Boulle; Diana Mulcahy; Annette Nell; René Toerien; Donovan Whitfield \par

\textbf{\large Siyavula and Free High School Science Texts contributors}\\

    Sarah Abel;
Dr. Rory Adams;
    Andrea Africa;
    Matthew Amundsen;
    Ben Anhalt;
    Prashant Arora;
    Amos Baloyi;
    Bongani Baloyi;
    Raymond Barbour;
    Caro-Joy Barendse;
    Richard Baxter;
    Tara Beckerling;
    Tim van Beek;
    Jennifer de Beyer;
Dr. Sarah Blyth;
    Sebastian Bodenstein;
    Martin Bongers;
    Gareth Boxall;
    Stephan Brandt;
    Hannes Breytenbach;
    Alex Briell;
    Wilbur Britz;
    Graeme Broster;
    Craig Brown;
    Deanne de Bude;
    Richard Burge;
    Bianca Böhmer;
    George Calder-Potts;
    Eleanor Cameron;
    Richard Case;
    Sithembile Cele;
    Alice Chang;
    Richard Cheng;
    Fanny Cherblanc;
Dr. Christine Chung;
    Brett Cocks;
    Stefaan Conradie;
    Rocco Coppejans;
    Tim Craib;
    Andrew Craig;
    Tim Crombie;
    Dan Crytser;
Dr. Anne Dabrowski;
    Laura Daniels;
    Gareth Davies;
    Sean Dobbs;
    Buhle Donga;
    William Donkin;
    Esmi Dreyer;
    Matthew Duddy;
    Fernando Durrell;
Dr. Dan Dwyer;
    Frans van Eeden;
    Alex Ellis;
    Tom Ellis;
    Andrew Fisher;
    Giovanni Franzoni;
    Ingrid von Glehn;
    Tamara von Glehn;
    Lindsay Glesener;
    Kevin Godby;
Dr. Vanessa Godfrey;
    Terence Goldberg;
Dr. Johan Gonzalez;
    Saaligha Gool;
    Hemant Gopal;
Dr. Stephanie Gould;
    Umeshree Govender;
    Heather Gray;
    Lynn Greeff;
    Carine Grobbelaar;
Dr. Tom Gutierrez;
    Brooke Haag;
    Kate Hadley;
    Alex Hall;
Dr. Sam Halliday;
    Asheena Hanuman;
Dr. Melanie Dymond Harper;
Dr. Nicholas Harrison;
    Neil Hart;
    Nicholas Hatcher;
    Jason Hayden;
    Laura Hayward;
Dr. William P. Heal;
    Pierre van Heerden;
Dr. Fritha Hennessy;
    Shaun Hewitson;
    Millie Hilgart;
    Grant Hillebrand;
    Nick Hobbs;
    Chris Holdsworth;
Dr. Benne Holwerda;
Dr. Mark Horner;
    Robert Hovden;
    Mfandaidza Hove;
    Jennifer Hsieh;
    Laura Huss;
    Rowan Jelley;
    Grant Jelley;
    Clare Johnson;
    Luke Jordan;
    Tana Joseph;
Dr. Fabian Jutz;
    Brian Kamanzi;
Dr. Lutz Kampmann;
    Simon Katende;
    Natalia Kavalenia;
    Nothando Khumalo;
    Paul Kim;
Dr. Jennifer Klay;
    Lara Kruger;
    Sihle Kubheka;
    Andrew Kubik;
Dr. Jannie Leach;
    Nkoana Lebaka;
Dr. Marco van Leeuwen;
Dr. Tom Leinster;
    Henry Liu;
    Christopher Loetscher;
    Mike Loseby;
    Amandla Mabona;
    Malothe Mabutho;
    Stuart Macdonald;
Dr. Anton Machacek;
    Tshepo Madisha;
    Batsirai Magunje;
Dr. Komal Maheshwari;
    Michael Malahe;
    Masoabi Malunga;
    Kosma von Maltitz;
    Masilo Mapaila;
    Bryony Martin;
    Nicole Masureik;
    John Mathew;
Dr. Will Matthews;
    Chiedza Matuso;
    JoEllen McBride;
    Nikolai Meures;
    Riana Meyer;
    Filippo Miatto;
    Jenny Miller;
    Abdul Mirza;
    Mapholo Modise;
    Carla Moerdyk;
    Tshwarelo Mohlala;
    Relebohile Molaoa;
    Marasi Monyau;
    Asogan Moodaly;
    Jothi Moodley;
    Robert Moon;
    Calvin Moore;
    Bhavani Morarjee;
    Kholofelo Moyaba;
    Nina Gitau Muchunu;
    Kate Murphy;
    Emmanuel Musonza;
    Tom Mutabazi;
    David Myburgh;
    Kamie Naidu;
    Nolene Naidu;
    Gokul Nair;
    Vafa Naraghi;
    Bridget Nash;
    Tyrone Negus;
    Huw Newton-Hill;
    Buntu Ngcebetsha;
Dr. Markus Oldenburg;
    Thomas O’Donnell;
Dr. Jaynie Padayachee;
    Poveshen Padayachee;
    Masimba Paradza;
    Dave Pawson;
    Justin Pead;
    Nicolette Pekeur;
    Sirika Pillay;
    Jacques Plaut;
    Barry Povey;
    Andrea Prinsloo;
    Joseph Raimondo;
    Sanya Rajani;
Prof. Sergey Rakityansky;
    Alastair Ramlakan;
Dr. Matina J. Rassias;
Dr. Jocelyn Read;
    Jonathan Reader;
    Jane Reddick;
Dr. Matthew Reece;
    Razvan Remsing;
    Laura Richter;
    Max Richter;
    Sean Riddle;
Dr. David Roberts;
    Christopher Roberts;
    Helen Robertson;
    Evan Robinson;
    Raoul Rontsch;
Dr. Andrew Rose;
    Katie Ross;
    Jeanne-Marié Roux;
    Mark Roux;
    Bianca Ruddy;
    Nitin Rughoonauth;
    Katie Russell;
    Steven Sam;
Dr. Carl Scheffler;
    Cho Hee Shrader;
    Nathaniel Schwartz;
    Duncan Scott;
    Helen Seals;
    Relebohile Sefako;
    Sandra Serumaga-Zake;
    Paul Shangase;
    Cameron Sharp;
    Ian Sherratt;
Dr. James Short;
    Roger Sieloff;
    Brandon Sim;
    Bonga Skozana;
    Clare Slotow;
    Bradley Smith;
    Greg Solomon;
    Nicholas Spaull;
Dr. Andrew Stacey;
Dr. Jim Stasheff;
    Mike Stay;
    Mike Stringer;
    Masixole Swartbooi;
    Tshenolo Tau;
    Tim Teatro;
    Ben Thompson;
    Shen Tian;
    Xolani Timbile;
    Nicola du Toit;
    Robert Torregrosa;
    Jimmy Tseng;
    Pieter Vergeer;
    Rizmari Versfeld;
    Mfundo Vezi;
    Mpilonhle Vilakazi;
    Mia de Vos;
    Helen Waugh;
    Leandra Webb;
Dr. Dawn Webber;
    Michelle Wen;
    Neels van der Westhuizen;
Dr. Alexander Wetzler;
Dr. Spencer Wheaton;
    Vivian White;
Dr. Gerald Wigger;
    Harry Wiggins;
    Heather Williams;
    Wendy Williams;
    Julie Wilson;
    Timothy Wilson;
    Andrew Wood;
    Emma Wormauld;
Dr. Sahal Yacoob;
    Jean Youssef;
    Ewald Zietsman; 
    Marina van Zyl





% Put the margins back for the rest of the book

\loadgeometry{fhsststandard}

\normalfont
